---
title: "Patterns and What Comes Next"
subtitle: "Find the rule, jump ahead"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 4, dpi = 150)
orange <- "#c2410c"
pink   <- "#f9c0c0"
blue   <- "#2563eb"
grey   <- "grey85"
```

class: center, middle

# Patterns

### Find the rule, and you can guess what comes next

(A grown-up walks through this with Isa. Lots of doing, not much reading.)

---

class: center, middle

# The big idea

When something follows a **pattern**, you can guess what comes next
**without building the whole thing**.

Finding that **rule** is the magic.

---

# Game 1: patterns that repeat

.pull-left[
Blue, red, blue, red, blue, red...

The grey ones are hidden.

**Ask her:** what comes next?
]

.pull-right[
```{r beads, fig.width=5, fig.height=1.6}
par(mar=c(0,0,0,0))
n <- 10; known <- 7
cols <- rep(c(blue, orange), length.out=n)
cols[(known+1):n] <- grey
plot(NA, xlim=c(0.5,n+0.5), ylim=c(-1,1), axes=FALSE, xlab="", ylab="")
points(1:n, rep(0,n), pch=21, bg=cols, col="grey40", cex=3, lwd=2)
text((known+1):n, 0, "?", col="grey50", cex=1.4)
```
]

---

class: center, middle

# What is the 10th one?

Blue, red, blue, red... every **even** spot is red.

So the 10th one is **red**, and you never had to point to each bead.

**Count by the repeat, not one at a time.**

---

# Game 2: patterns that grow

.pull-left[
Each step is a little **bigger** than the last.

These dots grow: **1, 3, 6, 10**.

**Ask her:** how many would the next one have?
]

.pull-right[
```{r grow, fig.width=5.5, fig.height=3}
par(mar=c(0,0,1,0))
plot(NA, xlim=c(-1,15), ylim=c(-0.5,4.5), axes=FALSE, xlab="", ylab="")
tri <- function(k, x0, col){
  for(r in 1:k){
    xs <- x0 + (seq_len(r) - (r+1)/2)
    points(xs, rep(k-r+1, r), pch=21, bg=col, col="grey40", cex=2, lwd=1.5)
  }
}
xs0 <- c(0, 3.5, 7.5, 12)
ks  <- c(1, 2, 3, 4)
for(i in 1:4) tri(ks[i], xs0[i], orange)
text(xs0, -0.2, c("1","3","6","10"), cex=1.4, font=2, col=orange)
```
]

---

class: center, middle

# Add a little more each time

Look at how many dots you **add**:

1 ... add **2** ... 3 ... add **3** ... 6 ... add **4** ... 10

So after 10 you add **5**. The next is **15**.

You found it without building it!

---

class: center, middle

# Another grower: doubling

**1, 2, 4, 8, ...**

Each step is **twice** the one before.

**Ask her:** what comes next? (Double 8 is **16**.) This one grows fast.

---

class: center, middle

# Every pattern has a rule

**Repeating** patterns come back to the start, over and over.

**Growing** patterns get bigger by a rule.

Once you know the rule, you can skip ahead.

---

# For the grown-up

- "How many would step 10 have?" is the seed of algebra. Keep it spoken and
  physical: build, guess, check. Do not push symbols.
- Repeating rows teach counting by groups. Growing rows teach that the *amount
  you add* can follow its own pattern (2, 3, 4, 5, ...).
- Rainy-day version: Mathigon Polypad has tiles and grids for building these.

---

class: center, middle

# Next time: number shapes

Those growing dots, **1, 3, 6, 10**, are real **triangles**.

In **Module 04** we build them, and the square numbers too.

That is the same triangle Ben proves things about. The play and the proof rhyme.
