---
title: "Mirror Magic"
subtitle: "Folds where both halves match"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 4, dpi = 150)
orange <- "#c2410c"
pink   <- "#f9c0c0"
```

class: center, middle

# Mirror Magic

### Some shapes have a secret fold

(A grown-up walks through this with Isa. Lots of doing, not much reading.)

---

class: center, middle

# The big idea

Some shapes have a **fold line** where one half lands exactly on the other half.

That line is called a **line of symmetry**.

---

# Game 1: the mirror finishes the picture

.pull-left[
- Draw **half** of something down the middle: half a heart, half a face.
- Stand a mirror on the middle line.
- The whole picture appears.

**Try letters!** Which ones does the mirror complete?
]

.pull-right[
### A  M  T  W
these work

### R  F  G
these do not
]

---

class: center, middle

# Game 2: fold and cut

Fold the paper. Cut a shape on the fold. Open it.

Both sides match. That fold is its line of symmetry.

```{r heart, fig.width=4.5, fig.height=3.6}
par(mar=c(0,0,0,0))
t <- seq(0, 2*pi, length.out=400)
x <- 16*sin(t)^3
y <- 13*cos(t) - 5*cos(2*t) - 2*cos(3*t) - cos(4*t)
plot(x, y, type="n", asp=1, axes=FALSE, xlab="", ylab="")
polygon(x, y, col=pink, border=orange, lwd=4)
abline(v=0, lty=2, col="grey40", lwd=2)
```

---

class: center, middle

# Game 3: the snowflake

Fold a circle of paper into a skinny wedge. Snip the edges. Open it slowly.

A real snowflake has **six** matching sides.

That is the same **six** that Ben builds with his compass!

---

# How many fold lines does a square have?

.pull-left[
Fold a square paper different ways. Each fold where the halves match is a line of
symmetry.

How many can you find?
]

.pull-right[
```{r square, fig.width=4.2, fig.height=4.2}
par(mar=c(0,0,0,0), pty="s")
plot(NA, xlim=c(-1.3,1.3), ylim=c(-1.3,1.3), asp=1, axes=FALSE, xlab="", ylab="")
rect(-1,-1,1,1, border=orange, lwd=4)
segments(0,-1, 0,1, lty=2, col="grey40", lwd=2)
segments(-1,0, 1,0, lty=2, col="grey40", lwd=2)
segments(-1,-1, 1,1, lty=2, col="grey40", lwd=2)
segments(-1,1, 1,-1, lty=2, col="grey40", lwd=2)
```
.center[The answer is **four**.]
]

---

class: center, middle

# Symmetry hunt

Look around the room. Find five things with a line of symmetry.

Faces, butterflies, leaves, letters, windows, forks.

---

# For the grown-up

- The whole idea is physical: "fold so the halves match." If she can fold a shape
  so the halves cover each other, it is symmetric; if not, it is not.
- Let her **find** that a square has four fold lines (two through the sides, two
  through the corners) by folding, not by being told.
- No-scissors rainy-day version: Mathigon Polypad has mirror and symmetry tools.
- This connects to Ben's compass hexagon: a snowflake's six-fold symmetry is the
  same six.
