---
title: "A Shape Hunt"
subtitle: "Find the whole course hiding in the real world"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 4, dpi = 150)
orange <- "#c2410c"
pink   <- "#f9c0c0"

# flat-top hexagon vertices around a center
hexpts <- function(cx, cy, s) {
  a <- (0:5) * pi/3
  list(x = cx + s*cos(a), y = cy + s*sin(a))
}
```

class: center, middle

# A Shape Hunt

### Go find the math in the real world

(A grown-up walks through this with Isa. Today we hunt, photograph, and sort.)

---

class: center, middle

# The big idea

We learned about **symmetry**, **shapes**, and **tilings**.

Today we go find them: in the house, in the yard, everywhere.

---

# Game 1: find these four shapes

```{r gallery, fig.width=9, fig.height=2.8}
par(mar=c(2,0,0,0))
plot(NA, xlim=c(0,12), ylim=c(-1.4,1.4), asp=1, axes=FALSE, xlab="", ylab="")
# circle
th <- seq(0, 2*pi, length.out=200)
polygon(1.5 + cos(th), sin(th), col=pink, border=orange, lwd=4)
text(1.5, -1.3, "circle", cex=1.4)
# triangle
polygon(c(4.5, 3.6, 5.4), c(1, -0.9, -0.9), col=pink, border=orange, lwd=4)
text(4.5, -1.3, "triangle", cex=1.4)
# rectangle
rect(6.7, -0.7, 8.6, 0.9, col=pink, border=orange, lwd=4)
text(7.65, -1.3, "rectangle", cex=1.4)
# hexagon
h <- hexpts(10.5, 0, 1)
polygon(h$x, h$y, col=pink, border=orange, lwd=4)
text(10.5, -1.3, "hexagon", cex=1.4)
```

Tally each one you spot. **Which shape did we find the most of?**

---

class: center, middle

# Hexagons are everywhere

A honeycomb is made of **hexagons**, six-sided shapes with no gaps.

```{r honeycomb, fig.width=7, fig.height=3.4}
par(mar=c(0,0,0,0))
plot(NA, xlim=c(-0.5,9), ylim=c(-0.5,4), asp=1, axes=FALSE, xlab="", ylab="")
s  <- 1
dx <- 1.5*s
dy <- sqrt(3)*s
for (col in 0:4) {
  for (row in 0:2) {
    cx <- col*dx
    cy <- row*dy + ifelse(col %% 2 == 1, dy/2, 0)
    h <- hexpts(cx, cy, s)
    polygon(h$x, h$y, col=pink, border=orange, lwd=3)
  }
}
```

The same **six** as the snowflake, and as Ben's compass hexagon!

---

class: center, middle

# Game 2: the symmetry hunt

Find things with a **fold line** where both halves match.

```{r butterfly, fig.width=5, fig.height=3.6}
par(mar=c(0,0,0,0))
ellipse <- function(cx, cy, a, b, n=120) {
  t <- seq(0, 2*pi, length.out=n)
  list(x = cx + a*cos(t), y = cy + b*sin(t))
}
plot(NA, xlim=c(-1.5,1.5), ylim=c(-1.4,1.4), asp=1, axes=FALSE, xlab="", ylab="")
# wings (mirror image left/right)
for (sgn in c(1, -1)) {
  up <- ellipse(sgn*0.6, 0.45, 0.55, 0.6)
  lo <- ellipse(sgn*0.5, -0.55, 0.45, 0.5)
  polygon(up$x, up$y, col=pink, border=orange, lwd=3)
  polygon(lo$x, lo$y, col=pink, border=orange, lwd=3)
}
# body
bd <- ellipse(0, 0, 0.08, 0.85)
polygon(bd$x, bd$y, col=orange, border=orange)
# fold line
abline(v=0, lty=2, col="grey40", lwd=2)
```

Faces, leaves, butterflies, forks, windows. **What is the most symmetric thing you can find?**

---

class: center, middle

# Game 3: the tiling hunt

Find a real **tiling**: a surface covered with no gaps.

A brick wall, a tile floor, a chessboard, a honeycomb.

```{r bricks, fig.width=7, fig.height=2.6}
par(mar=c(0,0,0,0))
plot(NA, xlim=c(0,8), ylim=c(0,3), asp=1, axes=FALSE, xlab="", ylab="")
for (row in 0:2) {
  off <- ifelse(row %% 2 == 1, -0.5, 0)
  for (i in -1:8) {
    x0 <- i*1 + off
    rect(x0, row, x0+1, row+1, col=pink, border=orange, lwd=3)
  }
}
rect(0, 0, 8, 3, border=NA)
```

These bricks are rectangles, with **no gaps**. That is the trick from Module 02.

---

class: center, middle

# Game 4: sort your finds

Photograph or draw what you found.

Sort them into piles: circles, triangles, rectangles, hexagons.

Put the symmetric ones together. Hang up your favorites.

---

class: center, middle

# You found the whole course!

**Symmetry** (Module 01) + **tilings** (Module 02) + the **hexagon** that links your snowflake to Ben's compass.

It was all hiding in plain sight.

### Now go play.

---

# For the grown-up

- This is a review and a celebration, not a test. No score, no right answers, a
  shrug is fine.
- Let the connections land lightly: the honeycomb hexagon ties back to the
  snowflake (Module 01) and to Ben's compass hexagon; the brick wall and tile
  floor are the tilings of Module 02.
- Keep the camera optional. Drawing the finds works just as well and slows the
  looking down.
- Rainy day indoors? Mathigon Polypad (see `../resources.md`) is full of shapes
  to find, sort, and tile on screen.
