---
title: "Number Shapes"
subtitle: "Numbers you can build out of dots"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 4, dpi = 150)
orange <- "#c2410c"
pink   <- "#f9c0c0"
blue     <- "#bfdbfe"
bluedark <- "#1d4ed8"

# n by n square of dots, lower-left corner at (x0, y0)
dot_square <- function(n, x0 = 0, y0 = 0, fill = pink, border = orange) {
  xs <- rep(0:(n - 1), times = n) + x0
  ys <- rep(0:(n - 1), each  = n) + y0
  points(xs, ys, pch = 21, bg = fill, col = border, cex = 2.4, lwd = 2)
}

# triangle number with rows 1..k, apex up, centered on x0, bottom row at y0
dot_triangle <- function(k, x0 = 0, y0 = 0, fill = pink, border = orange) {
  for (r in 1:k) {
    xs <- (seq_len(r) - (r + 1) / 2) + x0
    points(xs, rep(k - r, r) + y0, pch = 21, bg = fill, col = border,
           cex = 2.4, lwd = 2)
  }
}
```

class: center, middle

# Number Shapes

### Some numbers can be built out of dots

(A grown-up walks through this with Isa. Build the dots with real coins or blocks.)

---

class: center, middle

# The big idea

Some numbers have a **shape**.

You can build them out of dots, coins, or blocks, and the shape tells you
something about the number.

---

# Square numbers

```{r squares, fig.width=10, fig.height=3.2}
par(mar = c(0, 0, 0, 0))
plot(NA, xlim = c(-1, 17), ylim = c(-2, 4.5), asp = 1,
     axes = FALSE, xlab = "", ylab = "")
offs <- c(0, 2.5, 6, 10.5)
ns   <- 1:4
for (i in 1:4) dot_square(ns[i], x0 = offs[i], y0 = 0)
counts <- c(1, 4, 9, 16)
for (i in 1:4) text(offs[i] + (ns[i] - 1) / 2, -1.3, counts[i],
                    cex = 2, col = orange, font = 2)
```

**1, 4, 9, 16, 25**: each one is a full square with the same number on every side.

---

# Triangle numbers

```{r triangles, fig.width=10, fig.height=3.4}
par(mar = c(0, 0, 0, 0))
plot(NA, xlim = c(-1.5, 14), ylim = c(-2, 4), asp = 1,
     axes = FALSE, xlab = "", ylab = "")
offs <- c(0, 3, 7, 12)
ks   <- 1:4
for (i in 1:4) dot_triangle(ks[i], x0 = offs[i], y0 = 0)
counts <- c(1, 3, 6, 10)
for (i in 1:4) text(offs[i], -1.3, counts[i], cex = 2, col = orange, font = 2)
```

**1, 3, 6, 10, 15**: add one longer row each time (1, then 1+2, then 1+2+3 ...).

---

class: center, middle

# The surprise: two triangles make a rectangle

```{r rect, fig.width=6, fig.height=4.4}
par(mar = c(0, 0, 0, 0))
k <- 3; C <- k + 1
plot(NA, xlim = c(0.3, C + 0.7), ylim = c(-0.7, k - 0.3), asp = 1,
     axes = FALSE, xlab = "", ylab = "")
for (j in 1:k) for (i in 1:C) {
  if (i <= j) points(i, k - j, pch = 21, bg = pink, col = orange,
                     cex = 3, lwd = 2)
  else        points(i, k - j, pch = 21, bg = blue, col = bluedark,
                     cex = 3, lwd = 2)
}
```

Two **6**s lock into a **3 by 4** rectangle = 12. So a triangle number is **half a rectangle**.

---

# Squares grow by odd numbers

.pull-left[
Look at the square numbers in order:

1, then **+3** = 4, then **+5** = 9, then **+7** = 16.

Every new square adds an **L** of the next odd number of dots.
]

.pull-right[
```{r oddgrow, fig.width=4.4, fig.height=4.4}
par(mar = c(0, 0, 0, 0), pty = "s")
plot(NA, xlim = c(0.3, 4.7), ylim = c(0.3, 4.7), asp = 1,
     axes = FALSE, xlab = "", ylab = "")
for (i in 1:4) for (j in 1:4) {
  hl <- (i == 4 || j == 4)
  points(i, j, pch = 21, bg = ifelse(hl, orange, pink),
         col = orange, cex = 3, lwd = 2)
}
```
.center[The orange **L** is 7 dots: 9 + 7 = 16.]
]

---

class: center, middle

# Build them for real

Lay out coins or blocks. Make the squares. Make the triangles.

Fit two paper triangles together and watch the rectangle appear.

Ben builds the same **square numbers** as actual square tiles.

---

# For the grown-up

- The big moment is the rectangle: hand her two paper triangles (or two colors of
  counters) and let her **discover** that they fit, do not tell her first.
- The rule behind it: two copies of the n-th triangle number make an n by (n+1)
  rectangle. The picture *is* the proof; she does not need the formula.
- Cross-links: square dots connect to Ben's square tilings and to the pattern
  blocks of Module 02. The 1, 3, 6, 10 staircase is the one from Module 03.
- Rainy-day version: Mathigon Polypad has counters for building these on screen.

---

class: center, middle

# What's next

You found shapes hiding inside numbers.

Next: **Module 05, a shape hunt**. Go find these shapes out in the real world.
