---
title: "Pattern Blocks & Tangrams"
subtitle: "Shapes that fit together"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center", dev = "png",
                      fig.width = 5, fig.height = 4, dpi = 150)
orange <- "#c2410c"
pink   <- "#f9c0c0"
green  <- "#74b566"
```

class: center, middle

# Pattern Blocks & Tangrams

### Small shapes build big shapes

(A grown-up walks through this with Isa. Lots of doing, not much reading.)

---

class: center, middle

# The big idea

Shapes **fit together**.

You can build a big shape out of small ones, fill an outline more than one way,
and make pictures out of a few simple pieces.

---

# Fill the hexagon

.pull-left[
Take one yellow **hexagon**. Cover it exactly with green triangles.

How many fit?

**Six!**

Now try a different mix. How many ways can you fill the same hexagon?
]

.pull-right[
```{r hexagon, fig.width=4.2, fig.height=4.2}
par(mar=c(0,0,0,0), pty="s")
ang <- seq(0, 2*pi, length.out=7)
vx <- cos(ang); vy <- sin(ang)
plot(NA, xlim=c(-1.2,1.2), ylim=c(-1.2,1.2), asp=1, axes=FALSE, xlab="", ylab="")
for (i in 1:6) {
  polygon(c(0, vx[i], vx[i+1]), c(0, vy[i], vy[i+1]),
          col=green, border="white", lwd=3)
}
polygon(vx[1:6], vy[1:6], border=orange, lwd=5)
```
]

---

class: center, middle

# Tile the whole table

Cover the table edge to edge: **no gaps, no overlaps**.

```{r tiling, fig.width=7, fig.height=3.4}
par(mar=c(0,0,0,0))
hex <- function(cx, cy, r, col) {
  a <- (seq(0,5)*60 + 30) * pi/180
  polygon(cx + r*cos(a), cy + r*sin(a), col=col, border="white", lwd=2)
}
plot(NA, xlim=c(0,8), ylim=c(0,3.6), asp=1, axes=FALSE, xlab="", ylab="")
r <- 0.55; dx <- sqrt(3)*r; dy <- 1.5*r
cols <- c("#f6c453", green)
for (j in 0:4) {
  cy <- 0.5 + j*dy
  xoff <- if (j %% 2 == 0) 0 else dx/2
  for (i in 0:6) {
    cx <- 0.4 + xoff + i*dx
    hex(cx, cy, r, cols[(i+j) %% 2 + 1])
  }
}
```

That "no gaps" filling is a **tiling**. It is exactly what Ben studies!

---

class: center, middle

# The surprise: pentagons

Try to fit regular **pentagons** around a point.

```{r pentagons, fig.width=4.6, fig.height=4.6}
par(mar=c(0,0,0,0), pty="s")
pentagon <- function(alpha, col) {
  R <- 1
  beta <- c(180,108,36,324,252) * pi/180
  cx <- R*cos(alpha); cy <- R*sin(alpha)
  polygon(cx + R*cos(beta+alpha), cy + R*sin(beta+alpha),
          col=col, border="white", lwd=3)
}
plot(NA, xlim=c(-2.3,2.3), ylim=c(-2.3,2.3), asp=1, axes=FALSE, xlab="", ylab="")
pentagon(0,             "#6c8ebf")
pentagon(108*pi/180,    "#9673a6")
pentagon(216*pi/180,    "#b85450")
points(0, 0, pch=19, cex=0.7)
text(0.95*cos(288*pi/180), 0.95*sin(288*pi/180), "gap!",
     col=orange, font=2, cex=1.3)
```

They will **not** close up. There is always a gap.

---

# What can a tangram make?

.pull-left[
A **tangram** is one square cut into **7** pieces.

The one rule: every picture uses **all seven**, flat, no overlaps.

Put them back into the square. (Harder than it looks!)
]

.pull-right[
```{r tangram, fig.width=4.2, fig.height=4.2}
par(mar=c(0,0,0,0), pty="s")
plot(NA, xlim=c(-0.2,4.2), ylim=c(-0.2,4.2), asp=1, axes=FALSE, xlab="", ylab="")
P <- list(
  list(x=c(0,0,2),     y=c(0,4,2)),     # large
  list(x=c(0,4,2),     y=c(4,4,2)),     # large
  list(x=c(2,4,4),     y=c(2,4,2)),     # medium
  list(x=c(0,2,1),     y=c(0,0,1)),     # small
  list(x=c(1,2,3,2),   y=c(1,0,1,2)),   # square
  list(x=c(3,2,4),     y=c(1,2,2)),     # small
  list(x=c(2,4,4,3),   y=c(0,0,2,1))    # slanted piece
)
cols <- c("#6c8ebf","#82b366","#d6b656","#b85450",
          "#9673a6","#e8a87c","#76c1c4")
for (k in seq_along(P)) polygon(P[[k]]$x, P[[k]]$y, col=cols[k],
                                border="white", lwd=2.5)
rect(0,0,4,4, border=orange, lwd=5)
```
]

---

class: center, middle

# Make a picture

Use the pieces to build a **flower**, a **turtle**, a **rocket**, a **cat**.

No blocks? Cut a paper square into a tangram.

Then make **two different** pictures that each use all seven pieces.

---

# For the grown-up

- The richest question is **"how many different ways can you fill the hexagon."**
  Counting the ways is real combinatorial thinking dressed up as play.
- The **pentagons-leave-gaps** discovery is worth setting up on purpose. A regular
  pentagon's corner is 108 degrees, and 108 does not divide 360 evenly, so they
  cannot close around a point. You do not need to say that to Isa: let her find it.
- **No gaps** tiling is exactly Ben's tilings module, where he proves *why* only
  certain shapes tile.
- Rainy-day version: Mathigon Polypad has pattern blocks and tangrams on screen.

**Next:** Module 03, Patterns and what comes next.
